Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-08-27 fstats_pca_lewontin.tex
revision part 1

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies. Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical, but also on scientific grounds. In Lewontin's data, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002, reich2018a}, and we now know that human races are an arbitrary, biologically useless polyphyletic grouping that is maintained by historical and social conventions.

@@

This complex population structure is frequently handled by using multiple models with different assumptions; each emphasizing a particular aspects of the data. Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable. For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests.

@@

\subsection{Introduction to PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, there is no conceptual difference between an individual-based and population-based analysis. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains most of the variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	In many  implementations \citep[e.g.][]{patterson2006}, SNPs are weighted by
    the inverse of their standard deviations. As this weighting often makes little
    difference in practice \citep{mcvean2009}, I will assume throughout that SNPs
    are unweighted.

@@

\paragraph{``World'' data set}
This data set represents global human genetic variation (638 individuals from 33 population), as used by \citep{lazaridis2014}. As this data set is very sparse, it may be well-approximated by an admixture graph.

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is frequently used as a basis of comparison for ancient genetic analyses of Western Eurasian individuals \citep{patterson2012, lazaridis2014}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}, and thus may not be particularly graph-like.

\paragraph{Computing $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.

@@

\subsection{$F_4$-statistics as angles}
The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct. As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection (Figure \ref{fig:geom}E).

@@

\section{Discussion}
Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations). It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, and so higher PCs will be informative for population structure. Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment and sample sizes \citep{mcvean2009, novembre2008a}, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data

@@

An advantage of calculating $F$-statistics based on PCs is that they yield consistent estimtates. For both data sets I investigated here, the matrix $\MF$ of $F$-statistics obtained  using admixtools2 is not a proper squared Euclidean distance matrix, i.e. it is not negative semidefinite and has imaginary PCs. A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.
\newpage
\bibliographystyle{plainnat}
\bibliography{main}

\newpage
\appendix
\section{Derivations}\label{appendix:fonpc}
\setcounter{equation}{0}
\renewcommand{\theequation}{\thesection\arabic{equation}}
Depending on a readers' background in linear algebra, these results may appear elementary; I include them here for reference and because they were not obvious to me at the onset of this project.
\paragraph{$F$-statistics are invariant under a change-of-basis}
\begin{eqnarray}
F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
&=& \sum_k (P_{ik} - P_{jk})^2 \label{eq:changeofbasis}
\end{eqnarray}

>>>>

However, understanding, conceptualizing and modeling this variation is far from trivial, particularly since misconstrued models of human genetic differentiation have repeatedly been used to justify racist, eugenic and genocidal policies. Lewontin's landmark \citeyear{dobzhansky1972} paper on the apportionment of human genetic diversity was one of the first to question ``races'' not only on ethical grounds, but also based on genetic data. Based on 15 blood group loci, less than half of the already small proportion of between-population genetic variation can be attributed to ``racial'' continental-scale groupings. Over the last five decades, this view has been corroborated and extended many times \citep{cann1987, cavalli-sforza1994, cann2002, rosenberg2002a, reich2018a}, and we now know that human races cannot be motivated genetically, but rather are a polyphyletic grouping that is maintained by historical and social conventions.

@@

The interplay of these demographic processes leads to complex genetic structure that is difficult to model using standard population genetic models: Human populations are typically non-discrete, not at equilibrium and not randomly mating. One analysis paradigm is to use different tools based on different sets of assumptions; each emphasizing a particular aspects of the data. Data-driven methods such as Principal Component Analysis (PCA, \cite{cavalli-sforza1994}) structure \citep{pritchard2000} or multidimensional scaling (MDS, \cite{malaspinas2014}) are often used to display the full complexity of the data, but they have the disadvantage that they are not easily interpretable. For this purpose, more explicit demographic models \citep{gutenkunst2009, kamm2015, excoffier2013} are applied, which allow for parameter estimation or hypothesis tests, but with the drawback that strong assumptions about models need to be assumed.

@@

\subsection{Introduction to PCA}
Let us assume we have some genotype data summarized in a matrix $\MX$ whose entry $x_{ij}$ reflects the allele frequency of the $i$-th population at the $j$-th genotype. If we have $S$ SNPs and $n$ populations, $\MX$ will have dimension $n \times S$. As a population may be represented by just one individual, all results are valid if we treat each individual as its own population. Since the allele
frequencies are between zero and one, we can interpret each Population $X_i$
of $\MX$ as a point in $[0, 1]^S$, the allele frequency or \emph{data space}.

The goal of PCA is to find a low-dimensional subspace $\mathbb{R}^K$ that explains the most important variation in the data. $K$ is at most $n-1$, in which case the data is simply rotated. However, the  historical processes that generated genetic variation often result in \emph{sparse} data \citep{engelhardt2010}, so that $K \ll n$ explains a substantial portion of the variation; for visualization $K=2$ is frequently used.

@@

	There are a number of choices when constructing a PCA from genetic data, for example SNPs are sometimes standardized by the inverse of their standard deviations \citep{patterson2006}, or low-quality samples are projected onto PCs created from high-quality samples \citep{patterson2006}. In order to simplify the relationship between PCA and $F$-statistics, I assume no standardization and no missing data.

@@

\paragraph{``World'' data set}
This data set is a subset of the ``World Foci'' data set of \cite{lazaridis2014}, where I removed samples which are not permitted for free reuse. These populations span the globe and roughly represents global human genetic variation (638 individuals from 33 population) As this data set is very sparse, with often thousands of kilometers between adjacent sampling locations, I speculate that gene flow between these populations may not be particularly common; and their structure may therefore be well-approximated by an admixture graph. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 1.}

\paragraph{Western Eurasian data set}
This data set of 1,119 individuals from 62 populations contains  present-day individuals from the Eastern Mediterranean, Caucasus and Europe. It is motivated by the analysis of \cite{lazaridis2014}, who used it as a basis of comparison for ancient genetic analyses of Western Eurasian individuals, and PCAs based on similar sets of samples have been used in many other ancient DNA studies \citep[e.g.][]{lazaridis2016, haak2015}. Genetic differentiation in this region is low and closely mirrors geography \citep{novembre2008}. We thus might think that gene flow between these populations is common \citep{ralph2013}, and a discrete model such as a tree or an admixture graph might be a rather poor reflection of this data. A file with all individuals used and their assigned population is given in \textbf{Supplementary File 2.}

\paragraph{Computing $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs. A squared Euclidean distance matrix such as $\MF$ has all negative or zero eigenvalues (i.e. $\MF$ is negative-semidefinite). However as $F_2$-statistics are estimates, some eigenvalues might be slightly positive, which would lead to imaginary PCs. I avoid this by using the \texttt{nearPD}-function in R that ensures all eigenvalues have the correct sign.

@@

\subsection{$F_4$-statistics as angles}
The interpretation of $F_4$ in PCA is similar to that of $F_3$ as a projection of one vector onto another, with the difference that now all four points may be distinct. As for $F_3$, a finding of $F_4(X_1, X_2; X_3, X_4) = 0$ implies that the vectors $X_1 - X_2$ and $X_3 - X_4$ are orthogonal, or that the two populations map to the same point, and otherwise it will correspond to the length of the projection, multiplied by the distances $\norm{X_1 - X_2}$ and $ \norm{X_3 - X_4}$ (Figure \ref{fig:geom}E).

@@

\section{Discussion}
Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations). It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, higher PCs will be informative for population structure. Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment and sample sizes \citep{mcvean2009, novembre2008a}, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data

@@

An advantage of calculating $F$-statistics based on PCs is that they yield consistent estimtates. For both data sets I investigated here, the matrix $\MF$ of $F$-statistics obtained  using admixtools2 is not a proper squared Euclidean distance matrix, i.e. it is not negative semidefinite and has imaginary PCs. A model-based framework based on probabilistic PCA \citep{hastie2015, meisner2021, agrawal2020} would likely be able to generate consistent $F$-statistics and PCs, while incorporating sampling error and missing data.

\newpage
\bibliographystyle{plainnat}
\bibliography{main}

\newpage
\appendix
\section{Derivations}\label{appendix:fonpc}
\setcounter{equation}{0}
\renewcommand{\theequation}{\thesection\arabic{equation}}
Depending on a readers' background in linear algebra, these results may appear elementary; I include them here for reference and because they were not obvious to me at the onset of this project.
\paragraph{$F$-statistics are invariant under a change-of-basis}
\begin{eqnarray}
F_2(X_i, X_j) &=& \sum_{l=1}^S \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
&=& \sum_{l=1}^S \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{jk}\big)^2\nonumber\\
&=& \sum_{l=1}^S \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
&=& \sum_{l=1}^S \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^S L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
&=& \sum_k (P_{ik} - P_{jk})^2 \label{eq:changeofbasis}
\end{eqnarray}
